---
title: "Applied Linear Models Midterm"
date: "2023-10-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      cache = TRUE)
```

```{r data loading, message = FALSE, warning = FALSE}
library(tidyverse)
library(gridExtra)
library(car)
library(sf)
library(reshape2)
library(leaps)
library(glmnet)
load("Chicago.RData")
raw <- chicago

standardize <- function(v){
    return((v - mean(v, na.rm = TRUE)) / sd(v, na.rm = TRUE))
}

compute_mse <- function(model, validation, true_response){
    preds <- predict(model, newdata = validation)
    errors <- (preds - true_response)
    mse <- mean((errors ** 2))
    return(mse)
}
```

```{r}
# remove rows with 0 population
chicago <- chicago[!chicago$poptotal < 5, ]
# remove rows with NA for important predictors
chicago <- chicago[-which(is.na(chicago$income.male) | 
                              is.na(chicago$income.female)), ]

chicago$Ward <- factor(chicago$Ward)
chicago$Community.Area <- factor(chicago$Community.Area)
chicago$pctwhite <- round(chicago$popwhite / chicago$poptotal, digits = 3)
chicago$pctblack <- round(chicago$popblack / chicago$poptotal, digits = 3)
chicago$pctasian <- round(chicago$popasian / chicago$poptotal, digits = 3)
```

```{r transformations for q1}
cannabis_cidx <- startsWith(colnames(chicago), "CANNABIS")
noncannabis_cidx <- startsWith(colnames(chicago), "NOTCANNABIS")
cann_data <- chicago[, cannabis_cidx]
noncann_data <- chicago[, noncannabis_cidx]

chicago$cannabis_sum <- rowSums(chicago[, cannabis_cidx])
chicago$noncannabis_sum <- rowSums(chicago[, noncannabis_cidx])
chicago$crime_sum <- chicago$cannabis_sum + chicago$noncannabis_sum

chicago$crime_percap <- round(chicago$crime_sum / chicago$poptotal, digits = 3)
chicago$cannabis_percap <- round(chicago$cannabis_sum / chicago$poptotal, 
                                 digits = 3)
chicago$noncannabis_percap <- round(chicago$noncannabis_sum / chicago$poptotal, 
                                    digits = 3)

chicago$income.comb = chicago$income.male + chicago$income.female
```

```{r seasonal transformations}
# meteorological convention is summer is June, July, August
summer_cidx <- endsWith(colnames(chicago), "6") | 
    endsWith(colnames(chicago), "7") | endsWith(colnames(chicago), "8") 
winter_cidx <- endsWith(colnames(chicago), "12") | 
    endsWith(colnames(chicago), "1") | endsWith(colnames(chicago), "2")
spring_cidx <- endsWith(colnames(chicago), "3") | 
    endsWith(colnames(chicago), "4") | endsWith(colnames(chicago), "5")
fall_cidx <- endsWith(colnames(chicago), "9") | 
    endsWith(colnames(chicago), "10") | endsWith(colnames(chicago), "11")

chicago$summer_sum <- rowSums(chicago[, summer_cidx])
chicago$winter_sum <- rowSums(chicago[, winter_cidx])
chicago$spring_sum <- rowSums(chicago[, spring_cidx])
chicago$fall_sum <- rowSums(chicago[, fall_cidx])

chicago$summer_cann_sum <- rowSums(chicago[, summer_cidx & cannabis_cidx])
chicago$winter_cann_sum <- rowSums(chicago[, winter_cidx & cannabis_cidx])
chicago$spring_cann_sum <- rowSums(chicago[, spring_cidx & cannabis_cidx])
chicago$fall_cann_sum <- rowSums(chicago[, fall_cidx & cannabis_cidx])

chicago$summer_noncann_sum <- rowSums(chicago[, summer_cidx & noncannabis_cidx])
chicago$winter_noncann_sum <- rowSums(chicago[, winter_cidx & noncannabis_cidx])
chicago$spring_noncann_sum <- rowSums(chicago[, spring_cidx & noncannabis_cidx])
chicago$fall_noncann_sum <- rowSums(chicago[, fall_cidx & noncannabis_cidx])

chicago$summer_percap <- round(chicago$summer_sum / chicago$poptotal, 
                               digits = 3)
chicago$winter_percap <- round(chicago$winter_sum / chicago$poptotal, 
                               digits = 3)
chicago$spring_percap <- round(chicago$spring_sum / chicago$poptotal, 
                               digits = 3)
chicago$fall_percap <- round(chicago$fall_sum / chicago$poptotal, 
                             digits = 3)

chicago$summer_cann_percap <- round(chicago$summer_cann_sum / chicago$poptotal, 
                                    digits = 3)
chicago$winter_cann_percap <- round(chicago$winter_cann_sum / chicago$poptotal, 
                                    digits = 3)
chicago$spring_cann_percap <- round(chicago$spring_cann_sum / chicago$poptotal, 
                                    digits = 3)
chicago$fall_cann_percap <- round(chicago$fall_cann_sum / chicago$poptotal, 
                                  digits = 3)


chicago$summer_noncann_percap <- round(chicago$summer_noncann_sum / 
                                           chicago$poptotal, 
                                       digits = 3)
chicago$winter_noncann_percap <- round(chicago$winter_noncann_sum / 
                                           chicago$poptotal, 
                                       digits = 3)
chicago$spring_noncann_percap <- round(chicago$spring_cann_sum / 
                                           chicago$poptotal, 
                                       digits = 3)
chicago$fall_noncann_percap <- round(chicago$fall_noncann_sum / 
                                         chicago$poptotal, 
                                     digits = 3)


seasonal <- data.frame(percap = c(chicago$winter_percap, chicago$spring_percap, 
                                  chicago$summer_percap, chicago$fall_percap), 
                       season = c(rep("winter", nrow(chicago)), 
                                  rep("spring", nrow(chicago)), 
                                  rep("summer", nrow(chicago)), 
                                  rep("fall", nrow(chicago))), 
                       cann_percap = c(chicago$winter_cann_percap, 
                                       chicago$spring_cann_percap, 
                                       chicago$summer_cann_percap, 
                                       chicago$fall_cann_percap), 
                       noncann_percap = c(chicago$winter_noncann_percap, 
                                          chicago$spring_noncann_percap, 
                                          chicago$summer_noncann_percap, 
                                          chicago$fall_noncann_percap))
```


```{r final transformations}
modeldata <- chicago[, 1:13]
modeldata$pctwhite <- round(modeldata$popwhite / modeldata$poptotal, digits = 3)
modeldata$pctblack <- round(modeldata$popblack / modeldata$poptotal, digits = 3)
modeldata$pctasian <- round(modeldata$popasian / modeldata$poptotal, digits = 3)

melted <- melt(cann_data)
melted$code <- as.character(melted$variable)
tmp <- unlist(strsplit(melted$code, "[.]"))
melted$type = tmp[seq(1, length(tmp), 3)]
melted$year = tmp[seq(2, length(tmp), 3)]
melted$month = tmp[seq(3, length(tmp), 3)]
melted$variable <- NULL
melted$code <- NULL
melted$crime <- melted$value
melted$value <- NULL

cann_store <- melted

melted <- melt(noncann_data)
melted$code <- as.character(melted$variable)
tmp <- unlist(strsplit(melted$code, "[.]"))
melted$type = tmp[seq(1, length(tmp), 3)]
melted$year = tmp[seq(2, length(tmp), 3)]
melted$month = tmp[seq(3, length(tmp), 3)]
melted$variable <- NULL
melted$code <- NULL
melted$crime <- melted$value
melted$value <- NULL


# lengthen modeldata to match
modeldata <- modeldata %>% 
    slice(rep(1:n(), each = ncol(cann_data) + ncol(noncann_data)))

modeldata <- cbind(modeldata, rbind(cann_store, melted))

modeldata$season <- ifelse(modeldata$month %in% c(12, 1, 2), "winter", 
                           ifelse(modeldata$month %in% c(3, 4, 5), "spring", 
                                  ifelse(modeldata$month %in% c(6, 7, 8), 
                                         "summer", "fall")))
modeldata$season <- factor(modeldata$season)
modeldata$type <- factor(modeldata$type)

modeldata$crime_percap <- modeldata$crime / modeldata$poptotal
```

```{r model selection}
test_rows <- (modeldata$year == 2012 & modeldata$month %in% c(10, 11, 12))
validation <- modeldata[test_rows, ]

# fixing community area bug
validation[which(validation$Community.Area %in% c(31, 33, 34)), 
           "Community.Area"] <- 32

train <- modeldata[!test_rows, ]

selected <- regsubsets(crime_percap ~ income.male + income.female + 
                           factor(Community.Area) + factor(type) +
                           income.male + income.female +
                           pctblack:factor(type) + 
                           pctasian + pctwhite + factor(season), 
                         nvmax = 50,
                         method = "forward",
                         data = train)

summ <- summary(selected)
bic_min <- which(summ$bic == min(summ$bic))
selected <- colnames(summ$outmat)[which(summ$outmat[bic_min, ] == "*")]

model <- lm(crime_percap ~ income.male + factor(Community.Area) + 
                pctblack + pctwhite + type, 
            data = train)

mse <- compute_mse(model, validation, validation$crime_percap)
predictions <- predict(model, newdata = validation)

errors <- (predictions - validation$crime_percap) ** 2
# error needs to be transformed for the per captia adjustment
mse <- mean(errors * validation$poptotal)
```

```{r}
cat("The mean squared error of my predictor on the validation set is:", 
    round(mse, 5))

errorplt <- ggplot(data.frame(errors), aes(x = errors)) + 
    geom_histogram() + 
    labs(title = "Error distribution")

logerrplt <- ggplot(data.frame(errors), aes(x = log(errors))) + 
    geom_histogram() + 
    labs(title = "Error distribution", 
         subtitle = "Log transformed for readability")

grid.arrange(errorplt, logerrplt, nrow = 1)
```

\newpage

# Abstract

This report creates a predictor to predict the crime rate from October to December of 2012, using data from January 2010 to September of 2012. The final model relied heavily on income and community area as the primary variables of interest, and also retained as useful predictors the percentage of white and black inhabitants of the block. 

# Introduction

In this report we use criminal reports from the city of Chicago to graphically investigate whether and which block groups in Chicago have unusually high crime rates, how crime rates vary across Ward and Community Area, and how features like average income, season, gender distribution, and racial distribution interact and contribute to crime rates. Throughout the report we attempt to investigate whether cannabis crime and noncannabis crime rates have similar relationships to the data, and whether the relationships of the features to crime rate changes depending on whether we're predicting cannabis or noncannabis crime.

# Data

The data used were a collection of observations of criminal reports in a certain block area, with 2122 rows of data representing 2122 city blocks in city of Chicago. The key columns available in this data are summarized in the table below. However, extensive transformations were performed on the data set in order to better answer the questions of interest. In particular, we: 

1) Removed any rows with 0 population or `NA` values for important predictors (like income)
2) Converted crime totals to a per capita crime rate for each block, respecting the segmentation between cannabis and noncannabis crime rates. 
3) Computed racial percentages from the given data for each block
4) Compacted monthly data into seasonal per capita crime rates
5) Reformatted the data to be longer rather than wider, so each block did not appear as a single row but instead as a new row for each month. This means each row in the new dataset represents a "block-month(-year)" rather than just a single block.

| Variable(s)| Description |
| -----------|------------- |
| `poptotal`, `popwhite`, `popblack`, `popasian`| Count of ethnic group members in the FIPS group. |
| `income.male`, `income.female`| Average income in the FIPS group faceted by gender. |
| `age.male`, `age.female`| Average age of each gender in the FIPS group. |
| `longitude`, `latitude`| Longitude and latitude coordinates of the given FIPS group |
| `Ward`, `Community.Area`| Political/geographic demarcations within the city of Chicago. Ward is a voting district, Community Area changes less frequently. |
| Monthly crime totals (72 columns)| Total amounts of cannabis and noncannabis crime reported in the given block area. |

# Methods

To determine which variables were worth considering to predict crime rates we performed visual EDA as shown below on a transformed dataset After this, we included those variables in a forward selection cross validation to determine which variables were worth including in a final multiple linear regression model. We used a transformation to per capita crime rate (divided crime count by block population) through the paper in order to stablize the variance, but transformed back to volume units afterwards for the purpose of calculating the mean squared error.

# Results

## Identifying High Crime Areas

```{r, fig.height = 4}
chicago %>%
    ggplot(aes(x = longitude, y = latitude, 
               color = sqrt(crime_percap), 
               alpha = sqrt(crime_percap))) + 
    geom_point() + 
    scale_color_gradient(low = "lightblue", high = "darkorange") + 
    labs(title = "Geographical Distribution of Crime in Chicago", 
         x = "Longitude", y = "Latitude", 
         color = "Sqrt Crime Per Capita", 
         alpha = "Sqrt Crime Per Capita") + 
    theme_bw() + 
    theme(panel.grid = element_blank())
```

We can see clear "hotspot" criminal areas in this plot. However, neither latitude nor longitude are likely to be helpful **linear** predictors of that, especially when compared to information like Ward or Community Area, beacuse there are low crime areas both to the east, west, north, and south of each hotspot. However, this should incline us to suspect that capturing the outlier information will be valuable in minimizing prediction risk.

## Crime Rates by Ward and Community Area

Ward is a political delineation, Community Area is a geographical delineation. Ward changes much more frequently over the years than does Community Area.

```{r, fig.height = 3}
chicago %>%
    ggplot(aes(x = factor(Ward), y = log(crime_percap))) + 
    geom_violin(fill = "lightblue") + 
    geom_boxplot(fill = "darkblue", alpha = 0.5) + 
    labs(title = "Crime Rate by Ward", 
         x = "Ward", y = "Log Crime Rate") + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.text.x = element_blank(), 
          axis.ticks.x = element_blank())

chicago %>%
    ggplot(aes(x = factor(Community.Area), y = log(crime_percap))) + 
    geom_violin(fill = "lightblue") + 
    geom_boxplot(fill = "darkblue", alpha = 0.5) + 
    labs(title = "Crime Rate by Community Area",
         x = "Community Area", y = "Log Crime Rate") + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.text.x = element_blank(), 
          axis.ticks.x = element_blank())
```

We will disregard Ward information because it is a political boundary and often changed, while Community Area is more stable. We can also see from our graphs that Community Area captures more of the variation in the data than does Ward, since nearly all the boxplots for Ward cover overlap or are very near to overlapping, while the Community Area boxplots show much less overlap.

To determine whether there's a connection between cannabis and non-cannabis reports at the community area level, we'll take the average of each Community Area's cannabis and non-cannabis crime rates (per capita) and plot them against each other.

```{r, fig.height = 3}
chicago %>%
    group_by(Community.Area) %>%
    summarize(cannabis_avg = mean(cannabis_percap), 
              noncannabis_avg = mean(noncannabis_percap)) %>%
    ggplot(aes(x = cannabis_avg, y = noncannabis_avg)) + 
    geom_point() + 
    labs(title = "Community Area Avg Cannabis vs Noncannabis Crime Rate", 
         subtitle = "Average for each block in a community area",
         x = "Average Cannabis Crime Rate", 
         y = "Average Noncannabis Crime Rate") + 
    theme_bw() +
    theme(panel.grid = element_blank())
```

We see a very tight relationship between average cannabis crime rate and average noncannabis crime rate, even with a major outlier in both axes which follows the trend established by the rest of the data. In fact, by manual inspection of the data we can identify that the major outlier in this plot is the same as the brighest "hotspot" identified in the latitude/longitude plot of Chicago. This reinforces our inclination that cannabis and noncannabis crime are likely to follow the same behaviors, and that accounting for outliers by using Community Area is likely to be an effective way to predict that data.

## Correlation between Cannabis and Non-Cannabis crime

To determine whether cannabis and non-cannabis crime were related in the dataset, we plotted the two variables against each other. Adding features to control for race and income generally failed to improve the fit of the data; appendix (B) shows the relevant plots. Furthermore, we used partial F tests (see appendix C) to determine whether adding those same features and interaction terms significantly improved the performance of a linear model predicing noncannabis crime rate using cannabis crime rate, with generally null results. As you can see below there is a strong relationship between cannabis and noncannabis crime, indicating that the same methods used to predict one might be used to predict another. 

```{r, fig.height = 3}
ggplot(chicago, aes(x = log(cannabis_percap), y = log(noncannabis_percap))) + 
    geom_point(alpha = 0.2) + 
    labs(title = "Cannabis and Noncannabis crime are related", 
         x = "log Cannabis Crime Per Capita", 
         y = "log Noncannabis Crime Per Capita") + 
    theme_bw() +
    theme(panel.grid = element_blank())
```

## Investigating the effect of wealth and race

To investigate the effect of wealth and race on per capita crime rate, we plotted per capita crime rate against combined wealth (the summed income of both men and women on the relevant block) and against race percentage. Each block has three points plotted on the race relationship graph - one for each of the racial demographics tracked in the data - we did this because if we simply took the majority race in each block as representative we would lose a lot of potentially interesting information.

```{r, fig.height = 4}
income_plt <- ggplot(chicago, aes(x = income.comb, y = log(crime_percap))) + 
    geom_point(alpha = 0.2) + 
    labs(title = "Crime Rate by Combined Income", 
         subtitle = "Combined male and female income",
         x = "Combined Income", 
         y = "Log Crimes Per Capita") + 
    theme_bw() +
    theme(panel.grid = element_blank())

race <- data.frame(pct = c(chicago$pctwhite, chicago$pctasian, 
                           chicago$pctblack), 
                   race = c(rep("white", nrow(chicago)), 
                            rep("asian", nrow(chicago)), 
                            rep("black", nrow(chicago))), 
                   crime_percap = rep(chicago$crime_percap, 3))

race_plt <- ggplot(race, aes(x = pct, y = log(crime_percap), color = race)) + 
    geom_point(alpha = 0.3) + 
    labs(title = "Crime distribution by race",
         x = "Percentage of Race", 
         y = "Log Crime Rate") + 
    theme_bw() +
    theme(panel.grid = element_blank())

grid.arrange(income_plt, race_plt, nrow = 1)
```

In the combined income plot we see mostly no relationship, except for some very significant high outliers in low income blocks. In the crime rate by race plot we see that for the most part race does not seem to have a very consistent relationship with per capita crime rate, but similarly very large outliers on both the high and low ends appear to imply some use might come of including these variables in consideration for our final model.

## Investigating Seasonal Effects

In plotting the effect of season as below, we can see in most cases the seasonal effect is not very large, but persistent across both cannabis and noncannabis criminal activity. The effect, however, seems to run counter to the conventional wisdom in that the season with the most criminal activity in this data appears to be winter, rather than summer. This effect seems consistent enough in the data to be worth considering the effect of season in our final model.

```{r, fig.height = 4}
seasonal_plt <- ggplot(seasonal, aes(x = log(percap), fill = season)) +
    geom_density(alpha = 0.4) + 
    labs(title = "Crime Rate by Season",  
         x = "Log Per Capita Crime Rate", y = "Density") + 
    theme_bw() + 
    theme(panel.grid = element_blank())

seasonal_cann_plt <- ggplot(seasonal, aes(x = log(cann_percap), 
                                          fill = season)) +
    geom_density(alpha = 0.4) + 
    labs(title = "Cannabis Crime Rate by Season", 
         x = "Log Per Capita Crime Rate", y = "Density") + 
    theme_bw() + 
    theme(panel.grid = element_blank())

seasonal_noncann_plt <- ggplot(seasonal, aes(x = log(noncann_percap), 
                                             fill = season)) +
    geom_density(alpha = 0.4) + 
    labs(title = "Noncannabis Crime Rate by Season", 
         x = "Log Per Capita Crime Rate", y = "Density") + 
    theme_bw() + 
    theme(panel.grid = element_blank())

grid.arrange(seasonal_plt, seasonal_cann_plt, seasonal_noncann_plt, 
             nrow = 2)
```

## Final Model

Using forward selection, considering the effects of the variables `Community.Area`, `income.male`, `income.female`, `pctblack`, `pctwhite`, `pctasian`, `type`, and `season`, we settled on a linear regression model that retains `Community.Area`, `income.male`, `pctblack`, `pctwhite`, and `type` (type of crime predicted) as the relevant predictors.

To compute the MSE and errors histogram afterwards, each error was multiplied by the population total in its respective block to account for switching the model to predict per capita crime rate rather than the actual volume of crimes. The final model had a mean squared error of $0.005$ (after adjustment) on the validation (test) data set.

# Discussion

It is important to note that among the drawbacks of this model and approach are that we cannot make any causal inferences based on this analysis. Many of the variables that the model retained are likely extremely correlated to the number of police officers already in the area observing criminal activity, creating a "chicken and egg" problem - we can see a lot more crime in areas where we have police officers actively hunting for crime. 

With more time we would be interested in attempting to model this data again excluding the difference between cannabis and noncannabis crime, since the EDA mostly showed that the relationship between the two was strong.

# Appendices

## Appendix A: Identifying High Crime Rate Blocks

```{r}
tmp <- head(chicago[order(chicago$crime_percap, decreasing = TRUE), 
                    c("fips", "crime_sum", "crime_percap")], 10)
knitr::kable(tmp, 
             col.names = c("Block ID", "Absolute", 
                           "Per Capita"),
             caption = "Top 10 Crime-Heavy blocks")

tmp <- head(chicago[order(chicago$cannabis_percap, decreasing = TRUE), 
                    c("fips", "cannabis_sum", "cannabis_percap")], 10)
knitr::kable(tmp, 
             col.names = c("Block ID", "Absolute", 
                           "Per Capita"), 
             caption = "Top 10 Cannabis Crime-heavy blocks")

tmp <- head(chicago[order(chicago$noncannabis_percap, decreasing = TRUE), 
                    c("fips", "noncannabis_sum", "noncannabis_percap")], 10)
knitr::kable(tmp, 
             col.names = c("Block ID", "Absolute", 
                           "Per Capita"),
             caption = "Top 10 Non-Cannabis Crime-heavy blocks")
```

## Appendix B: Cannabis/Noncannabis relationship controlling for race and income

We can see in this plot faceted by majority race that the slope between cannabis crime rate and noncannabis crime rate is essentially identical across the plots, indicating that there is likely no effect.

```{r}
majority_race <- ifelse(chicago$popwhite > chicago$popblack &
                            chicago$popwhite > chicago$popasian, "white", 
                        ifelse(chicago$popblack > chicago$popasian, "black", 
                        "asian"))

ggplot(cbind(chicago, majority_race), 
       aes(x = log(cannabis_percap), y = log(noncannabis_percap))) + 
    geom_point(alpha = 0.2) + 
    labs(title = "Cannabis and Noncannabis crime are related", 
         x = "log Cannabis Crime Per Capita", 
         y = "log Noncannabis Crime Per Capita") + 
    theme_bw() +
    theme(panel.grid = element_blank()) + 
    facet_wrap(~majority_race)
```

## Appendix C: Anova outputs investigating model interactions

These outputs were what we used to investigate whether race or income were important factors to control for when characterizing the relationship between cannabis and noncannabis reports. The only model to show meaningful improvement on the null model (simply predicting noncannabis crime with cannabis crime) was the model with all interaction terms. This model is still significant with a bonferroni correction applied to the p-value.

```{r}
model3null <- lm(noncannabis_percap ~ cannabis_percap, data = chicago)

model3race <- lm(noncannabis_percap ~ cannabis_percap + pctwhite + pctasian + 
                     pctblack, data = chicago)

model3black <- lm(noncannabis_percap ~ cannabis_percap + pctblack, 
                  data = chicago)

model3income <- lm(noncannabis_percap ~ cannabis_percap + income.male + 
                       income.female, data = chicago)

model3main <- lm(noncannabis_percap ~ cannabis_percap + pctwhite + pctasian + 
                     pctblack + income.male + income.female, data = chicago)

model3interact <- lm(noncannabis_percap ~ cannabis_percap + 
                         cannabis_percap:income.male + 
                         cannabis_percap:income.female, 
                     data = chicago)

model3allint <- lm(noncannabis_percap ~ cannabis_percap + pctwhite:cannabis_percap 
                   + pctasian:cannabis_percap + pctblack:cannabis_percap + 
                       income.male:cannabis_percap + 
                       income.female:cannabis_percap, 
                   data = chicago)

anova(model3null, model3race)
anova(model3null, model3income)
anova(model3null, model3main)
anova(model3null, model3black)
anova(model3null, model3allint)
```