---
title: "2023 Score Distribution"
author: "Gabriel Krotkov"
date: "2023-03-26"
output: html_document
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(devtools)
library(tidyverse)

load_all("tbaR")
```

# Get Scout Comments 

```{r}
rm(list = ls())
path <- "kraken_apollo.csv"
data <- read.csv(path)

scouter <- "harold"

scouter_comments <- data[which(data$scoutName == scouter), 
                         c("teamNum", "comments")]

get_scouter_comments <- function(data, scouter){
    return(data[which(data$scoutName == scouter), 
                c("teamNum", "comments")])
}
```

# Exploration: Find Unscouted Matches

```{r}
data <- read.csv("krakendata/mainframe.csv")
schedule <- read.csv("krakendata/schedule.csv")

data_ids <- paste(data$teamNum, data$matchNum, sep = "-")
schedule_ids <- paste(schedule$red1, schedule$match_number, sep = " in q")
schedule_ids <- c(schedule_ids, 
                  paste(schedule$red2, schedule$match_number, sep = " in q"))
schedule_ids <- c(schedule_ids, 
                  paste(schedule$red3, schedule$match_number, sep = " in q"))
schedule_ids <- c(schedule_ids, 
                  paste(schedule$blue1, schedule$match_number, sep = " in q"))
schedule_ids <- c(schedule_ids, 
                  paste(schedule$blue2, schedule$match_number, sep = " in q"))
schedule_ids <- c(schedule_ids, 
                  paste(schedule$blue3, schedule$match_number, sep = " in q"))

remaining <- setdiff(schedule_ids, data_ids)
```

# Neat Function: Unscouted Matches

```{r}
library(reshape2)

data <- read.csv("krakendata/mainframe.csv")
schedule <- read.csv("krakendata/schedule.csv")

#' ID Unscouted
#' 
#' Identify and return all the team-match combinations that exist in schedule 
#' but don't exist in data.
#' @param data krakendata main frame, actual scouted data. Assumed to have a 
#' "matchNum" column and "teamNum" column with match and team number.
#' @param schedule csv schedule from Kraken, full match schedule. Assumed to 
#' have columns "match_number" and "red1", "red2", ... "blue1", ...
#' @param remove_unplayed (bool) TRUE to trim away matches from the schedule
#' if we haven't seen any matches that late or later.
id_unscouted <- function(data, schedule, remove_unplayed = TRUE){
    if (remove_unplayed){ # remove matches beyond your data
        match_limit <- max(data$matchNum, na.rm = TRUE)
        schedule <- schedule[schedule$match_number <= match_limit, ]
    }
    data_ids <- paste(data$teamNum, data$matchNum, sep = " in q")
    schedule <- schedule[, c("match_number", "red1", "red2", "red3", 
                             "blue1", "blue2", "blue3")]
    schedule <- melt(schedule, id = c("match_number"), 
                     value.name = "team", variable.name = "station")
    schedule$station <- NULL
    schedule_ids <- paste(schedule$team, schedule$match_number, sep = " in q")
    return(setdiff(schedule_ids, data_ids))
}

id_unscouted(data, schedule, TRUE)
```

# Generic Get Robot Field

```{r}
#' Indexer
#' 
#' @author Isolde Moyer
#' Helper function to access the matrix diagonal.
indexer <- function(row, col) {
    return(row[col])
}

#' Apply Indexer
#' 
#' Applies the indexer function to the dataframe and pulls out the diagonal 
#' of the resulting matrix, efficiently getting the requested data from DF.
#' @author Isolde Moyer
#' @details Assumes that the 'df' parameter is coercable to a matrix. Also
#' assumes that idx is a dataframe with columns "ridx" and "cidx" (standing for
#' row index and col index) that index into df. 
apply_indexer <- function(df, idx){
    indexed <- apply(as.matrix(df[idx$ridx, ]), 1, indexer, idx$cidx)
    mat <- do.call("cbind", indexed)
    return(diag(mat))
}

#' Substring Right
#' 
#' Returns the n rightmost characters in s.
#' @param s input string
#' @param n number of characters from the right to include (1 indexed)
substr_right <- function(s, n){
    return(substr(s, nchar(s) - n + 1, nchar(s)))
}

#' Inverted Substring Right
#' 
#' Returns string s without the n rightmost characters
#' @param s input string
#' @param n number of characters (1 indexed) to chop off the right.
substr_right_inv <- function(s, n){
    return(substr(s, 1, nchar(s) - 1))
}

#' Generic Robot Field Getter
#' 
#' generic solution for getting fields from tba that use "(color)_nameRobot#"
#' as a naming style. Assumes field does not have the leading underscore.
#' @param matches dataframe of match rows
#' @param field variable name of interest in the (color)_(field)Robot(#) format
#' @param team_id team id of interest
#' @param unlist (boolean) unlist the result? Vast majority of time TRUE, FALSE 
# if the content has complicated content not fit for a vector.
get_generic_robot_field <- function(matches, field_id, team_id, 
                                    unlist = TRUE){
    stations <- get_team_stations(matches, team_id)
    # assumption: station number is the last character of the string
    station_num <- substr_right(stations$station, 1)
    color <- substr_right_inv(stations$station, nchar(stations$station) - 1)
    cidx <- paste0(color, "_", field_id, "Robot", station_num)
    idx <- data.frame(ridx = stations$match, cidx = cidx)
    result <- apply_indexer(matches, idx)
    if (unlist) result <- unlist(result)
    return(result)
}
```

# Apply above script

I should automate some of the stuff I do here to take this "one level up" but I have to have some results for champs soon.

```{r}
team <- 316
fields <- c("mobility", "autoChargeStation", "endGameChargeStation")
matches <- team_matches(team, year = 2023)

# trim unplayed
matches <- matches[matches$blue_score != -1 & matches$red_score != -1, ]

mobility <- get_generic_robot_field(matches, fields[1], team)
auto_raw <- get_generic_robot_field(matches, fields[2], team)
endgame_raw <- get_generic_robot_field(matches, fields[3], team)

# account for docks/engages
stations <- get_team_stations(matches, team)
color <- substr_right_inv(stations$station, nchar(stations$station) - 1)
color <- color[order(stations$match)]
auto_level <- ifelse(color == "red", matches$red_autoBridgeState, 
                     matches$blue_autoBridgeState)
auto <- ifelse(auto_raw == "Docked", 
               ifelse(auto_level == "Level", "Engaged", "Docked"), 
               auto_raw)
endgame_level <- ifelse(color == "red", matches$red_endGameBridgeState, 
                        matches$blue_endGameBridgeState)
endgame <- ifelse(endgame_raw == "Docked", 
                  ifelse(endgame_level == "Level", "Engaged", "Docked"), 
                  endgame_raw)

table(mobility)
table(auto)
table(endgame)
```

This code chunk goes "one level up" but is not generalized.

```{r}
teams <- event_teams("2023mil", keys = TRUE)
fields <- c("mobility", "autoChargeStation", "endGameChargeStation")

get_tangibles <- function(team, fields, year = YEAR){
    matches <- team_matches(team, year = year)
    matches <- matches[matches$blue_score != -1 & matches$red_score != -1, ]
    
    # Needs generalization
    mobility <- get_generic_robot_field(matches, fields[1], team)
    auto_raw <- get_generic_robot_field(matches, fields[2], team)
    endgame_raw <- get_generic_robot_field(matches, fields[3], team)

    stations <- get_team_stations(matches, team)
    color <- substr_right_inv(stations$station, nchar(stations$station) - 1)
    color <- color[order(stations$match)]
    auto_level <- ifelse(color == "red", matches$red_autoBridgeState, 
                         matches$blue_autoBridgeState)
    auto <- ifelse(auto_raw == "Docked", 
                   ifelse(auto_level == "Level", "Engaged", "Docked"), 
                   auto_raw)
    endgame_level <- ifelse(color == "red", matches$red_endGameBridgeState, 
                            matches$blue_endGameBridgeState)
    endgame <- ifelse(endgame_raw == "Docked", 
                      ifelse(endgame_level == "Level", "Engaged", "Docked"), 
                      endgame_raw)
    return(list(mobility = table(mobility), auto = table(auto), 
                endgame = table(endgame)))
}

# yikes this takes a while to run
tangibles <- sapply(teams, get_tangibles, fields = fields)
```


# Identify Disagreements between Kraken and TBA

```{r}
# testing purposes - using txcmp1, no internet 
# DATA ONLY HAS QUAL MATCHES! THE SCRIPT WILL ASSUME IT ONLY GETS INPUT QUALS
tba_data <- event_matches("2023txcmp1", match_type = "qual")
save(tba_data, file = "apollo.rda")
```

```{r}
# define relevant cols
# mobility data, auto balance, endgame data

tba_cols <- c("blue1", "blue2", "blue3", "red1", "red2", "red3", "match_number",
                     "blue_autoChargeStationRobot1", 
                     "blue_autoChargeStationRobot2",
                     "blue_autoChargeStationRobot3", 
                     "blue_endGameChargeStationRobot1",
                     "blue_endGameChargeStationRobot2", 
                     "blue_endGameChargeStationRobot3",
                     "blue_mobilityRobot1", "blue_mobilityRobot2", 
                     "blue_mobilityRobot3", "red_autoChargeStationRobot1", 
                     "red_autoChargeStationRobot2",
                     "red_autoChargeStationRobot3", 
                     "red_endGameChargeStationRobot1",
                     "red_endGameChargeStationRobot2", 
                     "red_endGameChargeStationRobot3", 
                     "red_mobilityRobot1", "red_mobilityRobot2", 
                     "red_mobilityRobot3", "blue_autoBridgeState", 
                     "blue_endGameBridgeState", "red_autoBridgeState", 
                     "red_endGameBridgeState")

kraken_cols <- c("matchNum", "teamNum", "mobility", 
                 "autoBalance", "teleopBalance", "scoutName")

save(tba_cols, kraken_cols, file = "col_defs.rda")
```

```{r}
require(tidyverse)
# load and reformat data so the TBA data is kraken-compatible
load("col_defs.rda")
load("apollo.rda") # loads tba_data object
kraken_data <- read.csv("krakendata/kraken_apollo.csv")

kraken_data <- kraken_data %>%
  select(all_of(kraken_cols))
tba_data <- tba_data %>%
  select(all_of(tba_cols))

rm(tba_cols, kraken_cols)

col_stems <- c("", "_mobilityRobot", "_autoChargeStationRobot", 
               "_endGameChargeStationRobot")
col_names_template <- c("matchNum", "teamNum", 
                        "mobility", "autoBalance", "teleopBalance",
                        "autoBridgeState", "endgameBridgeState")

# pull out correct columns and rename columns
tba_blue1 <- tba_data[, c("match_number", paste0("blue", col_stems, "1"), 
                          "blue_autoBridgeState", "blue_endGameBridgeState")]
colnames(tba_blue1) <- col_names_template
tba_blue2 <- tba_data[, c("match_number", paste0("blue", col_stems, "2"), 
                          "blue_autoBridgeState", "blue_endGameBridgeState")]
colnames(tba_blue2) <- col_names_template
tba_blue3 <- tba_data[, c("match_number", paste0("blue", col_stems, "3"), 
                          "blue_autoBridgeState", "blue_endGameBridgeState")]
colnames(tba_blue3) <- col_names_template
tba_red1 <- tba_data[, c("match_number", paste0("red", col_stems, "1"), 
                         "red_autoBridgeState", "red_endGameBridgeState")]
colnames(tba_red1) <- col_names_template
tba_red2 <- tba_data[, c("match_number", paste0("red", col_stems, "2"), 
                         "red_autoBridgeState", "red_endGameBridgeState")]
colnames(tba_red2) <- col_names_template
tba_red3 <- tba_data[, c("match_number", paste0("red", col_stems, "3"), 
                         "red_autoBridgeState", "red_endGameBridgeState")]
colnames(tba_red3) <- col_names_template

tba_data <- rbind(tba_blue1, tba_blue2, tba_blue3, tba_red1, tba_red2, tba_red3)
rm(tba_red1, tba_red2, tba_red3, tba_blue1, tba_blue2, tba_blue3)
tba_data$teamNum <- as.integer(substr(tba_data$teamNum, 
                                      4, nchar(tba_data$teamNum)))
tba_data$mobility <- ifelse(tba_data$mobility == "Yes", TRUE, FALSE)

# conform balance data
tba_data$autoBalance <- ifelse(tba_data$autoBalance == "Docked", 
                               ifelse(tba_data$autoBridgeState == "Level", 
                                      "engage", "dock"), "na")
tba_data$teleopBalance <- ifelse(tba_data$teleopBalance == "Docked", 
                                 ifelse(tba_data$endgameBridgeState == "Level", 
                                        "engage", "dock"), 
                                 ifelse(tba_data$teleopBalance == "Park", 
                                        "park", "na"))
tba_data$autoBridgeState <- NULL; tba_data$endgameBridgeState <- NULL
kraken_data$teleopBalance <- ifelse(kraken_data$teleopBalance == "fail", 
                                    "na", kraken_data$teleopBalance)

# trim TBA data to remove matches beyond what we've scouted
match_limit <- max(kraken_data$matchNum)
tba_data <- tba_data[which(tba_data$matchNum <= match_limit), ]

# sort both dataframes
tba_data <- tba_data[order(tba_data$matchNum, tba_data$teamNum), ]
kraken_data <- kraken_data[order(kraken_data$matchNum, kraken_data$teamNum), ]
tba_data$id <- paste(tba_data$matchNum, tba_data$teamNum, sep = "-")
kraken_data$id <- paste(kraken_data$matchNum, kraken_data$teamNum, sep = "-")
```

```{r}
# find differences in the conformed data
# for each match ID in TBA, look to see if we have an equivalent in kraken
tba_ids <- paste(tba_data$id, tba_data$mobility, tba_data$autoBalance, 
                 tba_data$teleopBalance, sep = "-")
kraken_ids <- paste(kraken_data$id, kraken_data$mobility, 
                    kraken_data$autoBalance, kraken_data$teleopBalance, 
                    sep = "-")
unmatched <- which(!(tba_ids %in% kraken_ids))
tba_unmatched <- tba_data[unmatched, ]

# look only at unmatched for which we did at least scout the match
validation <- tba_unmatched[tba_unmatched$id %in% kraken_data$id, ]

```

challenge: write a code that, given a scout's name, will compute their % agreement with TBA

```{r}
scouted <- kraken_data
validation <- tba_data
scout_name <- "novinski"

```

``` {r}
nov_vec <- scouted$scoutName == "novinski"
nov_df <- scouted[nov_vec, 1:7]
nov_validation <- validation[which(validation$id %in% nov_df$id), 1:6]
nov_df <- nov_df[which(nov_df$id %in% nov_validation$id), ]

colnames(nov_validation) <- paste0(colnames(nov_validation), "_validation")
nov_combined <- cbind(nov_df, nov_validation)
nov_combined$mobility_agree <- nov_combined$mobility == nov_combined$mobility_validation
sum(nov_combined$mobility_agree) / nrow(nov_combined)
```

```{r}
agreement_mobility <- function(scout_name, scouted, validation){
    scout_df <- scouted[scouted$scoutName == scout_name, ]
    scout_validation <- validation[which(validation$id %in% scout_df$id), ]
    scout_df <- validation[which(scout_df$id %in% scout_validation$id), ]
    colnames(scout_validation) <- paste0(colnames(scout_validation), 
                                         "_validation")
    scout_combined <- cbind(scout_df, scout_validation)
    tmp <- scout_combined$mobility == scout_combined$mobility_validation
    scout_combined$mobility_agree <- tmp
    return(sum(scout_combined$mobility_agree) / nrow(scout_combined))
}

agreement_generic <- function(scout_name, col_name, scouted, validation){
    scout_df <- scouted[scouted$scoutName == scout_name, ]
    scout_validation <- validation[which(validation$id %in% scout_df$id), ]
    scout_df <- validation[which(scout_df$id %in% scout_validation$id), ]
    colnames(scout_validation) <- paste0(colnames(scout_validation), 
                                         "_validation")
    scout_combined <- cbind(scout_df, scout_validation)
    scout_combined$agree <- scout_combined[, col_name] == 
      scout_combined[, paste(col_name, "validation", sep = "_")]
    return(sum(scout_combined$agree) / nrow(scout_combined))
}

scouted$scoutName <- tolower(scouted$scoutName) # match character cases
scouted$scoutName <- str_trim(scouted$scoutName) # remove whitespace around edges
scouted$scoutName <- ifelse(scouted$scoutName == "henry horvat",
                       	"horvat", scouted$scoutName)
scouts <- unique(scouted$scoutName)

mobility_pcts <- rep(0, length(scouts))
auto_balance_pcts <- rep(0, length(scouts))
endgame_pcts <- rep(0, length(scouts))
for (i in 1:length(scouts)){
  mobility_pcts[i] <- agreement_generic(scouts[i], "mobility", 
                                        scouted, validation)
  auto_balance_pcts[i] <- agreement_generic(scouts[i], "autoBalance", 
                                            scouted, validation)
  endgame_pcts[i] <- agreement_generic(scouts[i], "teleopBalance", 
                                       scouted, validation)
}
names(mobility_pcts) <- scouts
names(auto_balance_pcts) <- scouts
names(endgame_pcts) <- scouts
```


# Access Scouter Comments

```{r}
path <- "krakendata/mainframe.csv"
data <- read.csv(path)
data$scoutName <- tolower(data$scoutName)

scout_counts <- table(data$scoutName)
scouters <- names(scout_counts)

scouter_comments <- data[which(data$scoutName == scouter), 
                         c("teamNum", "comments")]

get_scouter_comments <- function(data, scouter){
    return(data[which(data$scoutName == scouter), 
                c("teamNum", "comments")])
}

for (i in 1:length(scouters)){
    comments <- get_scouter_comments(data, scouters[i])
    write.csv(comments, file = paste0("comments/", scouters[i], ".csv"))
}
```

# SuperScout helper

```{r}
# get links & score, output to a csv
# REQUIREMENT: needs to be able to see a tbaR package in its local directory
# OUTPUT: "ssframe.csv", in whatever local directory we're in

library(devtools)
load_all("tbaR")

event <- "2023mil"
matches <- event_matches("2023mil", match_type = "qual")
# filter out unplayed matches
matches <- matches[(matches$red_score != -1) & (matches$blue_score != -1), ]

# we can use match_number since we only requested qual data from tbaR
result <- data.frame(matchNum = matches$match_number, 
                     redScore = matches$red_score, 
                     redLinks = matches$red_linkPoints / 5,
                     blueScore = matches$blue_score, 
                     blueLinks = matches$blue_linkPoints / 5)
result <- result[order(result$matchNum), ]
write.csv(result, file = "ssframe.csv")
```

