---
title: "36-401 Data Exam 1"
author: "Jane Doe"
date: "October 13, 2023"
output: pdf_document
fontsize: 12pt
geometry:  margin = 1 in
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
# By default, do not include R source code in the PDF. We do not want to see
# code, only your text and figures.
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message=FALSE}
library(ggplot2)
library(dplyr)
library(broom)
```

```{r}
sleep_data = read.csv("/Users/Jane/Desktop/CMU/Year 4/401/Data Exam 1/cmu-sleep.csv")
sleep_data = sleep_data %>%
  mutate(TotalSleepTime = TotalSleepTime/60)
```

# Introduction

Carnegie Mellon University and other universities have recently collected data about
their students sleep habits, their academic performance and commitment, as well as
other demographic data. Also, CMU has changed their course scheduling to allow 8 am
courses due to classroom space restrictions as well as other registrar scheduling.
As a result, though, both students and professors have complained due to low attendance,
seemingly lower class performance, and less sleep. Previous studies have shown that
lack of sleep is correlated with poor academic performance so CMU's registrar's office
has tasked us with investigating the relationship between nightly sleep and GPA,
as a measure of student performance. 

This relationship will be investigated using the collected data to determine if there
is evidence for an association between sleep time and GPA, what the expected change, 
on average, in GPA is for a student who gets two few hours of sleep, and if the 
change in GPA can be proven to be caused by changes in sleep. The interest in a 2 hour
sleep change is because that has been estimated to be the effective sleep loss for
waking up earlier than normal for an 8 am class.

In the end, we used a linear regression model and found a positive association between
GPA and sleep which is consistent with the worries of the registrar's office. And, based
on evidence of how the data was collected, concluded a non-causal relationship. While
this model had some drawbacks stemming from skewed variables and a lack of randomly 
assigned treatment groups, we still found some evidence for the negative affects 
a lack of sleep has on academic performance which should discourage 8 am class usage.


# Exploratory Data Analysis & Data Summary

This data set contains information from 634 students across three universities who were investigating
the effect of 8 am classes on a student's sleep and thus their performance in class. The participants 
wore sleep trackers (Fitbits) for a month which recorded the average time students slept each night. 
This study occurred in the spring semester and performance of the students was evaluated via GPA. 
Both the GPA of the student during the semester of the study and the prior semester are included in 
the data set as two separate variables. While the study also collected other information from the 
students such as demographics, how long they napped for, and how many credit units they were taking
in the semester to name a few, the primary variables of interest to us are listed below.

* **TotalSleepTime**: The average time in hours the student spent asleep each night, not counting naps.
Originally collected in minutes but converted to hours for ease of interpretability.

* **term_gpa**: The student's GPA (out of 4) for the classes they took in the semester being studied (spring).

* **cum_gpa**: The student's GPA (out of 4) for the semesters before the one being studied (fall).


### Data Visualization

First, we will examine these three variables on their own.  

```{r, message = FALSE, fig.width=4, fig.height=3, fig.cap="Histogram of time students spent sleeping at night."}
ggplot(sleep_data, aes(x = TotalSleepTime)) +
  geom_histogram() +
  labs(x = "Time spent sleeping at night (hours)",
       y = "Count",
       title = "Sleep Time Histogram")
```

The histogram of total sleep time shown in Figure 1 appears roughly symmetrical and 
Normal-like with no strong skews. It is centered around 7 hours or 400 minutes of 
sleep per night. Most of the data points are concentrated between 4 and 9 hours with
some more extreme points around 3  and 10 hours of sleep.

```{r, message = FALSE, fig.width=4, fig.height=3, fig.cap="Histogram of student's GPA the semester the study occured."}
ggplot(sleep_data, aes(x = term_gpa)) +
  geom_histogram() +
  labs(x = "Student's term GPA (out of 4)",
       y = "Count",
       title = "Term GPA Histogram")
```

```{r, message = FALSE, fig.width=4, fig.height=3, fig.cap="Histogram of student's GPA prior to the study occuring."}
ggplot(sleep_data, aes(x = cum_gpa)) +
  geom_histogram() +
  labs(x = "Student's cumulative GPA (out of 4)",
       y = "Count",
       title = "Cumulative GPA Histogram")
```

The histograms of both the term and cumulative GPA variables are shown in Figures 2 
and 3 respectively. They are both strongly skewed left as student's GPAs tend to be 
concentrated between 3 and 4, closer to 4. There are also some potential outliers
with GPAs less than 0.5 but they seem few and should not cause many problems going
forward. One would normally want their response and or predictor variables less extreme, 
so transformations such as `log(constant-x)` could be used to better future modeling.

```{r, message = FALSE, fig.width=4, fig.height=3, fig.cap="Scatterplot of term GPA vs. sleep time."}
ggplot(sleep_data, aes(y = term_gpa, x = TotalSleepTime)) +
  geom_point() +
  labs(y = "Student's term GPA (out of 4)",
       x = "Nightly sleep time (hours)",
       title = "Term GPA vs. Sleep Time")
```

```{r, message = FALSE, fig.width=4, fig.height=3, fig.cap="Scatterplot of cumulative GPA vs. sleep time."}
ggplot(sleep_data, aes(y = cum_gpa, x = TotalSleepTime)) +
  geom_point() +
  labs(y = "Student's cumulative GPA (out of 4)",
       x = "Nightly sleep time (hours)",
       title = "Cumulative GPA vs. Sleep Time")
```

Figure 4 shows the relationship between term GPA and sleep time. It is roughly linear 
and positive. Figure 5 shows the relationship between cumulative GPA and sleep time. 
This is also roughly linear and positive but appears to be a bit of a weaker relationship than 
term GPA vs. sleep time. In both of these scatter plots, most of the data points are 
concentrated between 6 to 8 hours and close to 4 GPA units which is to be expected from
the univariate analysis. This is due to the strong skew of the GPA variables and concentration
of sleep time around 7 hours which makes it harder to visually examine the data. 

```{r, message = FALSE, fig.width=4, fig.height=3, fig.cap="Scatterplot of term GPA vs. cumulative GPA."}
ggplot(sleep_data, aes(y = term_gpa, x = cum_gpa)) +
  geom_point() +
  labs(y = "Student's term GPA (out of 4)",
       x = "Student's cumulative GPA (out of 4)",
       title = "Term GPA vs. Cumulative GPA")
```

Figure 6 shows a strong, positive linear relationship between the two GPA terms with
most of the data points concentrated between (3,3) and (4,4). This relationship makes
intuitive sense as student performance, on average, does not probably change that
drastically within one year.

There are some limitations to this data as well. It was not collected via randomly
assigned treatment nor were certain possible relevant variables polled such as university
prestige, economic background, or others. As a result, the conclusions made in this 
report can not easily be transferred or generalized to other universities that might be
very different from CMU. Also, the data visualization shown above proves weakly 
linear relationships between sleep and either GPA which would normally indicate 
variable transformation or non-linear modeling. 


# Methods

As we are looking to examine how student performance or GPA changes as a function of 
time slept at night. We can fit a simple linear regression that predicts term GPA from
total time slept at night. This is a valid model to make as a potential linear
relationship can be seen between these two variables as shown in Figure 4 above.
This model was made in terms of term GPA not cumulative GPA because, visually, 
there appeared to be a stronger linear relationship when comparing Figures 
4 and 5. It also makes more intuitive and practical sense to relate the sleep time to GPA for the 
semester in which the sleep data was gathered which in this case would be term.

The equation of the regression will follow this:

`term_GPA` = $\beta_0$ + `TotalSleepTime`$\beta_1$ + $\epsilon_i$

This model will help us answer the questions asked of this report as an association
between these two variables, term GPA and sleep time, can be evaluated via $\hat{\beta_1}$.
This number represents the slope of the predicted linear relationship between the two
variables so if we can be reasonably confident that it is not statistically significantly
different from zero, there is evidence for an association which would help us to answer
the registrar's question.


```{r, results='hide'}
#fitting model
term.model = lm(term_gpa~TotalSleepTime, data = sleep_data)
```

Before using this described model to answer the report questions, we need to check
diagnostics to have confidence in the validity of the numbers and conclusions. First,
we will examine the model to check for influential points via Cook's Distance. 

```{r, results='hide', include = FALSE}
#cook's distance
cookd.term = cooks.distance(term.model)
which.max(cookd.term)

head(augment(term.model))

plot = augment(term.model) %>%
  ggplot(aes(x = TotalSleepTime, y = .cooksd)) +
  geom_point() + 
  labs(y = "Cook's Distance",
       x = "Total Nightly Sleep Time (hours)",
       title = "Cook's Distance vs. Sleep Time")

sort(pf(cookd.term,2,632), decreasing = TRUE) #none of the points have cooks distances above the 50th percentile of f distribution with df 2 and n-2 so none are influential enough to remove 
curve(df(x, df1=2, df2=632))
```

```{r, message = FALSE, fig.width=4, fig.height=3, fig.cap="Plot of the Cook's Distance for Term GPA Model."}
plot
```

Figure 7 displays the Cook's Distances for each data point in the model as a function
of nightly sleep. Most of these values are quite small with one point, 471, being the 
most influential with a Cook's Distance of 0.204. To evaluate, though, if a Cook's 
distance value is too large and thus removal should be considered, one can compare
the values to the quantiles of the F distribution. In this case, none of the points
have Cook's Distances above the 50th percentile of a F distribution with n = 2 and 
n-2 = 632 degrees of freedom, so are not of concern. 

Next, we will examine the residuals of the model to check the linear regression 
requirements that the mean of the residuals is zero and that the spread is constant.
Figure 8 displays the residual or difference between predicted GPA and true GPA for 
each data point as a function of nightly sleep. This plot shows that the constant
spread and mean of zero are somewhat justified. The data points do seem to be 
concentrated around zero with roughly the same number above and below y = 0 but 
there is still a pattern to the plot with the line that is formed toward the top. 
There is not a strict triangular pattern to the plot which would indicate non-constant
variance which is good. 

```{r}
#plot of residuals v sleep time
plotr = augment(term.model) %>%
  ggplot(aes(x = `TotalSleepTime`, y = .resid)) +
  geom_point() +
  labs(y = "Residuals",
       x = "Total Nightly Sleep Time (hours)",
       title = "Residuals vs. Sleep Time")
```

```{r, message = FALSE, fig.width=4, fig.height=3, fig.cap="Plot of the Residuals for Term GPA Model."}
plotr
```

The next diagnostic we will look at is a QQ plot to check if the residuals appear 
to be Normally distributed. While this is an important aspect to consider when
choosing and evaluating a model, the large sample size of the data we are examining
means that the distribution of $\hat{\beta_1}$ will be approximately Normal even if 
the residuals are not due to the Central Limit Theorem. 

```{r}
#qq plot
plotqq =
  ggplot(augment(term.model), aes(sample = .resid)) +
  geom_qq() +
  geom_qq_line() +
  labs(y = "Standardized Residuals",
       x = "Theoretical Quantiles",
       title = "Normal Q-Q Plot")
```

```{r, message = FALSE, fig.width=4, fig.height=3, fig.cap="QQ Plot for Term GPA Model."}
plotqq
```

Figure 9 shows this QQ plot for our model. There is variation from a line at the
extremes of the theoretical quantiles indicating that the residuals are only 
approximately Normal at best. But as mentioned above, the sample size is so large 
that the distributions of the beta estimates will be approximately Normal regardless 
of the distribution of the residuals so we can trust hypothesis tests and confidence 
intervals performed on these estimates. 

Overall, the error assumptions of a mean of zero and constant spread and a roughly
linear relationship between sleep time and term GPA allows us to use this model to 
estimate the association between our two variables of interest and make predictions.
As mentioned during the data visualization section and looking at the residual plot in
Figure 8, the strong skew of the GPA variable is likely to be causing some issues 
in regard to the strength of the linear relationship and the error assumptions of 
linear regression. This is a limitation of our model. A log transformation like mentioned
during the EDA section of the report was tried to correct the strong skew of GPA but 
the improvement to randomness in the residual plot did not seem significant enough to 
continue with the model compared to the great decrease in interpretability as a 
result of the transformation. One could have also considered adding additional predictor
variables to the model but again, interpretability would decrease and possible problems
with multicolinearity could arise. It is important to note, though, that due to some of
the linear regression assumptions only weakly being met, especially linearity and the
residuals having a mean of zero, the model might not as accurately predict the true
relationship between term GPA and sleep as if more model investigation had been done. 

Now, we will use this model to answer the research questions by performing a hypothesis
test on $\beta_1$ and finding a confidence interval for the estimate produced by the 
linear regression model. In order to perform these tests, the linear regression
assumptions must be met which we proved above and the distribution of $\hat{\beta_1}$
must be approximately Normal which we know is the case due to the large sample
size. The hypothesis test will be set up in the following way.

* Ho: $\beta_1$ = 0
* Ha: $\beta_1$ $\not$= 0
* Test Statistic: T ~ t(n-2) where T = $\hat{\beta_1}$ / SE($\hat{\beta_1}$) 
* Rejection Region: T > $t_{1-\alpha/2, n-2}$ OR T < -$t_{1-\alpha/2, n-2}$ where $\alpha$ = 0.05 and n = 634


# Results

The model described above which predicts term GPA from total hours of sleep provided
results which are outlined below: 

```{r, results='hide'}
summary(term.model)
```

Betas                            Estimate   Std. Error  t value   p value
-------------------------------  ---------  ----------  --------  --------
$\hat{\beta_0}$ - Intercept        2.6610    0.1536      17.327   < 2e-16
$\hat{\beta_1}$ - TotalSleepTime   0.1191    0.0230      5.176    3.04e-07 

Table 1: Summary of results from linear regression model. 


The $R^{2}$ was found to be 0.04067 meaning that roughly 4% of the variation in 
term GPA is explained by the model. Also, the residual standard error of the model
was 0.4906 based on n-2 = 632 degrees of freedom.

The results of the hypothesis test described in the methods section are also shown 
in Table 1 showing that the test statistic, T, is 5.176 and the corresponding 
p value is 3.04e-07. This is proof that according to this model, there is evidence 
that term GPA is associated with hours of sleep per night (t(632) = 5.176, p<3.04e-07). 

```{r, results='hide'}
confint(term.model, "TotalSleepTime", 0.95)
2*confint(term.model, "TotalSleepTime", 0.95)
```

On average, students with one more hour of sleep per night have a term GPA that is 
$\hat{\beta_1}$ = 0.12 units higher (95% CI[0.0739, 0.164]). This confidence interval 
is above zero indicating that as hours of sleep increases, so does term GPA. As such,
there is evidence that students who sleep less might get lower term GPAs. 

One of the research questions is to predict if a student were to sleep 2 hours less,
on average, what average GPA effect do we expect to see. This relates to the 
linear regression equation, more specifically $\hat{\beta_1}$. The estimate of the slope
represents the change in GPA if hours were to change by one. If you multiply the slope
by 2, you get how much GPA will change if students sleep changes by 2. So, if a student
were to sleep 2 hours less, on average, their GPA should decrease by 0.24 units 
(95% CI[0.148, 0.329]). In some instances, a GPA drop of that size could lower students
from a B to C average which might greatly impact their confidence, moral, or future job/school
prospects.

The final research question of the report asked if we can conclude that the 
GPA change is caused by less sleep, or if their could be other reasons for the 
relationship. Since this data was gathered as an observational study meaning 
students were not randomly assigned to a trial group, we can not conclude a causal 
relationship. For example, had the study been performed where students were assigned
to either attend or not attend an 8 am class and then GPA was recorded, one would
be able to make causal claims as to if attending an 8 am class affected GPA had they
held confounding variables constant. In this instance, sleep was was just observed 
as well as GPA. Even just this data set has 12 other variables that were not focused 
on and these could have influenced a student's GPA just as much as their sleep. 

# Conclusion

As a result of the data analysis and modeling performed, there is evidence that
less sleep might harm academic performance which is normally something that a university
like Carnegie Mellon should care a lot about. We found through performing a linear
regression model that association between sleep and GPA is a positive one and that 
losing 2 hours of sleep per night on average could detriment a student's GPA by almost 
0.25 units. As a result, we would recommend trying to find alternatives to the scheduling 
challenges than using the 8 am time slot. While this is our recommendation, we can not 
conclude that lack of sleep is causing a lower GPA just that there is a correlation 
between these two variables. The data set provided included many other variables 
such as demographics, napping times, and term units all of which could have influenced 
term GPA as well. There are even other variables not measured that could have an effect 
such as university student culture, economic background, or major. This could be a 
place for future work such that a stronger, more accurate relationship of the many 
factors influencing term GPA could be established. Another place for improvement 
could come in the modeling. This model was based on only one variable and the strong 
skew which is intrinsic to a variable like GPA could benefit from future transformations 
that were deemed to difficult to interpret in this context. As such, while this data 
set nor the modeling was perfect, there was some evidence that this relationship 
could exist such that future digging should be done both for the students' and 
professors' well being and success going forward.




